---
title: "tximport_deseq2"
output: html_document
date: "2025-07-17"
params:
  quant_dir: "out/salmon"           # directory containing per-sample Salmon dirs with quant.sf
  gtf_file: "data/references/gtf/Mus_musculus.GRCm39.109.gtf"  # path to GTF (default now under data/references)
  project_name: "test"              # project prefix for all output filenames
  install_pkgs: false                # set TRUE to allow install during render (use conda env instead)
  tximport_rds: !r NULL              # optional precomputed tximport object (.rds)
  sample_table: "CU25_sample_info.csv"  # sample table CSV (rows = samples)
  group_col: "trx"                  # column in sample_table used for design
  out_dir: "out/deseq2"             # output directory for figures/tables
  padj_thresh: 0.05                  # adjusted p-value threshold for summaries/volcano
  lfc_thresh: 0.5                    # absolute log2FC threshold for volcano highlighting
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Preparing Data for DESeq2 with tximport and labeling Gene IDs
```{r}
# -----------------------------
# 1. Load libraries (installs optional; prefer conda env)
# -----------------------------
if (isTRUE(params$install_pkgs)) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
  BiocManager::install(c("tximport","GenomicFeatures","txdbmaker"), ask = FALSE, update = FALSE)
}

suppressPackageStartupMessages({
  library(tximport)
  library(readr)
  library(GenomicFeatures)
  library(txdbmaker)
  library(dplyr)
  library(tibble)
})

# Parameters (backward-compatible defaults in YAML)
quant_dir    <- params$quant_dir
gtf_file     <- params$gtf_file
out_dir      <- params$out_dir
padj_thresh  <- params$padj_thresh
lfc_thresh   <- params$lfc_thresh
tximport_rds <- params$tximport_rds
project_name <- params$project_name
file_prefix  <- paste0(project_name, "_")

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# -----------------------------
# 2. Define file paths 
# -----------------------------
# Update this to the actual path of your GTF and quant directories
# (now parameterized via YAML params above)
# gtf_file <- "reference/Mus_musculus.GRCm39.109.gtf"  # kept default via params
quant_dirs <- list.dirs(quant_dir, recursive = FALSE, full.names = TRUE)

# -----------------------------
# 3. Create transcript → gene map from GTF
# -----------------------------
txdb <- makeTxDbFromGFF(gtf_file, format = "gtf")
k <- keys(txdb, keytype = "TXNAME")
tx2gene <- AnnotationDbi :: select(txdb, 
                                   keys = k, 
                                   columns = c("GENEID"), 
                                   keytype = "TXNAME")

# Optional: convert to clean tibble
tx2gene <- as_tibble(tx2gene)

# -----------------------------
# 4. Read Salmon quant.sf files (or load provided tximport RDS)
# -----------------------------
if (!is.null(tximport_rds)) {
  txi <- readRDS(tximport_rds)
} else {
  # Each folder in quant_dirs should contain a quant.sf file
  files <- file.path(quant_dirs, "quant.sf")
  names(files) <- basename(quant_dirs)

  # Import transcript-level Salmon quantification, summarize to gene-level
  txi <- tximport(files, 
                  type = "salmon", 
                  tx2gene = tx2gene,
                  ignoreTxVersion = TRUE)
}

# -----------------------------
# 5. Add Gene Symbols (Optional)
# -----------------------------
if (isTRUE(params$install_pkgs) && !requireNamespace("biomaRt", quietly = TRUE)) {
 install.packages("biomaRt", repos = "https://cloud.r-project.org")
}
suppressPackageStartupMessages(library(biomaRt))

# Strip version suffixes from Ensembl IDs (ENSG00000123456.5 -> ENSG00000123456)
gene_ids_raw <- rownames(txi$counts)
gene_ids <- sub("\\..*$", "", gene_ids_raw)

# Auto-detect species from GTF filename
dataset_name <- if (grepl("Homo_sapiens|GRCh", gtf_file, ignore.case = TRUE)) {
  "hsapiens_gene_ensembl"
} else if (grepl("Mus_musculus|GRCm", gtf_file, ignore.case = TRUE)) {
  "mmusculus_gene_ensembl"
} else {
  warning("[Rmd] Cannot auto-detect species from GTF filename. Defaulting to human.")
  "hsapiens_gene_ensembl"
}

message("[Rmd] Using biomaRt dataset: ", dataset_name)

annot <- tryCatch({
  message("Fetching gene annotations from Ensembl biomaRt...")
  ensembl <- useEnsembl(biomart = "genes", dataset = dataset_name)
  result <- getBM(
    attributes = c("ensembl_gene_id", "external_gene_name"),
    filters = "ensembl_gene_id",
    values = unique(gene_ids),
    mart = ensembl
  )
  # Remove duplicate Ensembl IDs (keep first occurrence)
  result <- dplyr::distinct(result, ensembl_gene_id, .keep_all = TRUE)
  message("Retrieved annotations for ", nrow(result), " genes")
  result
}, error = function(e) {
  warning("biomaRt annotation failed: ", conditionMessage(e),
          "\nProceeding without gene symbol annotation.")
  data.frame(
    ensembl_gene_id = unique(gene_ids),
    external_gene_name = NA_character_,
    stringsAsFactors = FALSE
  )
})

# Force sane types
annot <- as.data.frame(annot)
annot$ensembl_gene_id  <- as.character(annot$ensembl_gene_id)
annot$external_gene_name <- as.character(annot$external_gene_name)

# Merge gene symbols (use stripped gene IDs for matching)
counts_with_symbols <- as.data.frame(txi$counts)
counts_with_symbols$ensembl_gene_id <- sub("\\..*$", "", rownames(counts_with_symbols))
counts_with_symbols$ensembl_gene_id <- as.character(counts_with_symbols$ensembl_gene_id)

counts_annotated <- dplyr::left_join(
 counts_with_symbols,
 annot,
 by = "ensembl_gene_id"
)

counts_annotated <- counts_annotated %>%
 dplyr::select(external_gene_name, dplyr::everything(), -ensembl_gene_id) %>%
 dplyr::rename(gene_symbol = external_gene_name)

# -----------------------------
# 6. Save outputs
# -----------------------------

# Rename columns to explicit sample IDs from the sample table
sample_table_path <- params$sample_table
st <- read.csv(sample_table_path, row.names = 1, check.names = FALSE)
sample_ids <- rownames(st)

clean_names_from_sample_table <- function(current_names, sample_ids) {
  escape_regex <- function(x) gsub("([][{}()+*^$|\\?.])", "\\\\\\1", x)
  vapply(current_names, function(nm) {
    hits <- sample_ids[sapply(sample_ids, function(s) {
      pat <- paste0("(^|[^A-Za-z0-9])", escape_regex(s), "($|[^A-Za-z0-9])")
      grepl(pat, nm, ignore.case = TRUE, perl = TRUE)
    })]
    if (length(hits) == 1) hits else nm
  }, character(1))
}

# Apply to txi counts and annotated counts
colnames(txi$counts) <- clean_names_from_sample_table(colnames(txi$counts), sample_ids)

library(dplyr)

counts_annotated <- counts_annotated %>%
  dplyr::select(gene_symbol, dplyr::everything())

count_cols <- setdiff(colnames(counts_annotated), "gene_symbol")
new_count_cols <- clean_names_from_sample_table(count_cols, sample_ids)

# Diagnostics: report columns that did not map uniquely
diagnose_mapping <- function(current_names, sample_ids) {
  escape_regex <- function(x) gsub("([][{}()+*^$|\\?.])", "\\\\\\1", x)
  res <- lapply(current_names, function(nm) {
    hits <- sample_ids[sapply(sample_ids, function(s) {
      pat <- paste0("(^|[^A-Za-z0-9])", escape_regex(s), "($|[^A-Za-z0-9])")
      grepl(pat, nm, ignore.case = TRUE, perl = TRUE)
    })]
    list(name = nm, hits = hits)
  })
  none <- vapply(res, function(x) length(x$hits) == 0, logical(1))
  multi <- vapply(res, function(x) length(x$hits) > 1, logical(1))
  if (any(none)) {
    message("[Rmd] Sample name mapping: no unique match for: ", paste(vapply(res[none], `[[`, character(1), "name"), collapse = ", "))
  }
  if (any(multi)) {
    details <- vapply(res[multi], function(x) paste0(x$name, " -> {", paste(x$hits, collapse = "/"), "}"), character(1))
    message("[Rmd] Sample name mapping: multiple matches: ", paste(details, collapse = "; "))
  }
}

diagnose_mapping(count_cols, sample_ids)
colnames(counts_annotated)[match(count_cols, colnames(counts_annotated))] <- new_count_cols

# Reorder to match sample table first, then any remaining
ordered_cols <- c("gene_symbol",
                  sample_ids[sample_ids %in% new_count_cols],
                  setdiff(new_count_cols, sample_ids))
counts_annotated <- counts_annotated[, ordered_cols, drop = FALSE]

# Export raw counts for reference (NOT used for DESeq2 analysis)
# These CSV files are provided for user reference only
# DESeq2 analysis uses the tximport object directly to preserve gene-length offsets
write.csv(counts_annotated, file = file.path(out_dir, paste0(file_prefix, "gene_counts_with_symbols.csv")), row.names = FALSE)
write.csv(txi$counts, file = file.path(out_dir, paste0(file_prefix, "raw_gene_counts.csv")))

cat("✔ Gene-level counts saved as:\n- ", file.path(out_dir, paste0(file_prefix, "gene_counts_with_symbols.csv")), "\n- ", file.path(out_dir, paste0(file_prefix, "raw_gene_counts.csv")), "\n", sep = "")
```


# Loading packages

```{r}

if (isTRUE(params$install_pkgs)) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
      install.packages("BiocManager", repos = "https://cloud.r-project.org")
  BiocManager::install("DESeq2", ask = FALSE, update = FALSE)
  if (!requireNamespace("tidyverse", quietly = TRUE))
    install.packages("tidyverse", repos = "https://cloud.r-project.org")
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(dplyr)
})

```

# Prepare sample metadata for DESeq2

```{r}
# Load sample data
colData <- read.csv(params$sample_table, row.names = 1)

# Verify that sample names in txi match sample names in colData
# The txi column names should already match the sample_ids from the renaming above
txi_samples <- colnames(txi$counts)
coldata_samples <- rownames(colData)

# Check for mismatches
missing_from_coldata <- setdiff(txi_samples, coldata_samples)
missing_from_txi <- setdiff(coldata_samples, txi_samples)

if (length(missing_from_coldata) > 0 || length(missing_from_txi) > 0) {
  stop("Sample name mismatch between Salmon quantifications and sample_table!\n",
       if (length(missing_from_coldata) > 0)
         paste0("  In Salmon but not in sample_table: ",
                paste(missing_from_coldata, collapse = ", "), "\n"),
       if (length(missing_from_txi) > 0)
         paste0("  In sample_table but not in Salmon: ",
                paste(missing_from_txi, collapse = ", "), "\n"),
       "Ensure sample_table rownames exactly match Salmon output folder names.")
}

# Reorder colData to match txi (important for DESeq2)
colData <- colData[txi_samples, , drop = FALSE]

# Verify alignment
message("Samples aligned: ", paste(txi_samples, collapse = ", "))
all(colnames(txi$counts) == rownames(colData))
```


# Contructing a DESeqDataSet object

```{r}
message("Creating DESeq2 dataset directly from tximport object (preserves length offsets)")

design_formula <- as.formula(paste("~", params$group_col))
dds <- DESeqDataSetFromTximport(txi,
                                 colData = colData,
                                 design = design_formula)
dds

# pre-filtering: removing rows with low gene counts
# keeping rows that have at least 10 reads total
dds <- dds[ rowSums(counts(dds)) >= 10, ]

# set the factor level
dds[[params$group_col]] <- relevel(dds[[params$group_col]], ref = "control")
```

#NOTE: if you have technical replicates collapse them
```

# Run DESeq

```{r}
dds <- DESeq(dds)

res <- results(dds)
res

# baseMean = average of normalized counts
# stat = Waldtest stats

# saving DEGs to data table for further analysis and saving data_table to csv
res_df <- as.data.frame(res)

# Add gene symbols to results (merge annotation created earlier)
# Strip version suffixes from Ensembl IDs for matching
res_df$ensembl_gene_id <- sub("\\..*$", "", rownames(res_df))
res_df <- dplyr::left_join(res_df, annot, by = "ensembl_gene_id")
res_df <- res_df %>%
  dplyr::select(ensembl_gene_id, external_gene_name, dplyr::everything())

write.csv(res_df,
          file = file.path(out_dir, paste0(file_prefix, "DESeq2_full_results.csv")),
          row.names = FALSE)
```

# Explore results

```{r}
summary(res)
# upregulated/downregulated genes, outliers. Default padj is .1

# increasing cutoff to < params$padj_thresh
res0.05 <- results(dds, alpha = padj_thresh)
summary(res0.05)

# contrasts
resultsNames(dds)

# MA plot
plotMA(dds)
```

# Volcano plot

```{r}
# Make sure you have ggrepel installed
if (isTRUE(params$install_pkgs) && !requireNamespace("ggrepel", quietly=TRUE)) {
  install.packages("ggrepel", repos = "https://cloud.r-project.org")
}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggrepel)
})

# Clamp padj for -log10 transformation (avoid Inf from padj = 0)
res_df$padj_plot <- ifelse(
  is.na(res_df$padj),
  NA_real_,
  pmax(res_df$padj, .Machine$double.xmin)
)

# Filter out NA values for the plot
volcano_data <- res_df %>% filter(!is.na(padj_plot))

# Add column to specify up vs downregulated
volcano_data$diffexpressed <- "NO"
volcano_data$diffexpressed[volcano_data$log2FoldChange >  lfc_thresh &
                            volcano_data$padj < padj_thresh] <- "Up"
volcano_data$diffexpressed[volcano_data$log2FoldChange < -lfc_thresh &
                            volcano_data$padj < padj_thresh] <- "Down"

# Gene labels (top 30) - use gene symbols if available, otherwise Ensembl ID
volcano_data$delabel <- NA
top_genes <- head(volcano_data[order(volcano_data$padj), ], 30)
for (i in seq_len(nrow(top_genes))) {
  if (!is.na(top_genes$external_gene_name[i]) &&
      top_genes$external_gene_name[i] != "") {
    volcano_data$delabel[which(rownames(volcano_data) ==
                                rownames(top_genes)[i])] <-
      top_genes$external_gene_name[i]
  } else {
    volcano_data$delabel[which(rownames(volcano_data) ==
                                rownames(top_genes)[i])] <-
      top_genes$ensembl_gene_id[i]
  }
}

volcano_plot <- ggplot(data = volcano_data,
                       aes(x = log2FoldChange, y = -log10(padj_plot),
                           col = diffexpressed, label = delabel)) +
  geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), col = "gray", linetype = "dashed")+
  geom_hline(yintercept = -log10(padj_thresh), col = "gray", linetype = "dashed")+
  geom_point(size =.5)+
  scale_color_manual(values = c("#00AFBB", "grey", "red"),
                     labels = c("Downregulated", "Not Significant", "Upregulated")) +
  labs(color = 'Up vs Down', x = expression("log"[2]*"FC"), y = expression("-log"[10]*"padj")) +
  ggtitle("Control vs ID8 (p53-/-)") +
  geom_text_repel(max.overlaps = Inf)

ggsave(file.path(out_dir, paste0(file_prefix, "volcano_plot.pdf")), 
       plot = volcano_plot,
       device = pdf,
       width = 10,
       height = 8)
```

# VST normalized count files and PCA plot

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = params$group_col)

# Add gene symbols to VST normalized counts
vst_counts <- as.data.frame(assay(vsd))
vst_counts$ensembl_gene_id <- rownames(vst_counts)
vst_counts <- dplyr::left_join(vst_counts, annot, by = "ensembl_gene_id")
vst_counts <- vst_counts %>%
  dplyr::select(ensembl_gene_id, external_gene_name, dplyr::everything())

write.csv(vst_counts,
          file = file.path(out_dir, paste0(file_prefix, "vst_norm_counts.csv")),
          row.names = FALSE)

pca_plot <- plotPCA(vsd, intgroup = params$group_col) + stat_ellipse(linetype = 3)

ggsave(file.path(out_dir, paste0(file_prefix, "PCA_plot.pdf")),
       plot = pca_plot,
       device = pdf,
       width = 7,
       height = 5)
```

# Enricher

```{r}

# libraries
suppressPackageStartupMessages({
  library(enrichR)
  library(writexl)
  library(openxlsx)
})

# Defining Databases
db <- c("KEGG_2019_Mouse", "Reactome_2022", "GO_Biological_Process_2025", "GO_Cellular_Component_2025", "GO_Molecular_Function_2025", "Tabula_Muris", "Elsevier_Pathway_Collection", "Bioplanet_2019")

# Subsetting DEGs to only upregulated using only top 200 genes
goi <- res_df %>%
  dplyr::filter(padj < padj_thresh, log2FoldChange > 1.5) %>%
  dplyr::slice_min(order_by = padj, n = 200)

# Extracting gene list as character vector (use symbols if available, otherwise Ensembl IDs)
genes <- ifelse(!is.na(goi$external_gene_name) & goi$external_gene_name != "",
                goi$external_gene_name,
                goi$ensembl_gene_id)
genes <- as.character(genes)

# Enrichment
edb<- enrichr(genes, db)


# Creating Excel workbook
wb <-openxlsx::createWorkbook()

for (db_name in db){
  addWorksheet(wb, sheetName = db_name)
  writeData(wb,
            sheet = db_name,
            x = edb[[db_name]],
            rowNames = FALSE)
}

# saving workbook
saveWorkbook(wb,
             file = file.path(out_dir, paste0(file_prefix, "Upregulated_cachexia_Enrichment_results.xlsx")),
             overwrite = TRUE)


# Subsetting DEGs to only downregulated using only top 200 genes
goi <- res_df %>%
  dplyr::filter(padj < padj_thresh, log2FoldChange < -1.5) %>%
  dplyr::slice_min(order_by = padj, n = 200)

# Extracting gene list as character vector (use symbols if available, otherwise Ensembl IDs)
genes <- ifelse(!is.na(goi$external_gene_name) & goi$external_gene_name != "",
                goi$external_gene_name,
                goi$ensembl_gene_id)
genes <- as.character(genes)

# Enrichment
edb<- enrichr(genes, db)


# Creating Excel workbook
wb <-openxlsx::createWorkbook()

for (db_name in db){
  addWorksheet(wb, sheetName = db_name)
  writeData(wb,
            sheet = db_name,
            x = edb[[db_name]],
            rowNames = FALSE)
}

# saving workbook
saveWorkbook(wb,
             file = file.path(out_dir, paste0(file_prefix, "Downregulated_cachexia_Enrichment_results.xlsx")),
             overwrite = TRUE)

```
